A Simplified Plackett-Luce Likelihood Function for Rank Orders

### 2024-08-16 inst/hankin-AustrianJournalofStatistics.tex
typos and tweaks

Ranked data arises whenever multiple entities are sorted by some
criterion; the preferred interpretation is a race in which the
sufficient statistic is the order of competitors crossing the
finishing line.  The Plackett--Luce likelihood
function~\cite{luce1959,plackett1975} is a generalization of the
Bradley--Terry model~\cite{bradley1952} in which non-negative strengths
are assigned to each competitor; without loss of generality, if the
finishing order is $1,2,\ldots,n$ and competitor $i$ has strength
$p_i\geqslant 0$, then the Plackett--Luce likelihood function will be
proportional to

@@

\noindent where the strengths are normalised so $\sum p_i=1$.
However, estimation of the strengths is somewhat difficult if $n$ is
even moderately large: even with as few as 20 competitors following
Zipf's law~\cite{zipf1949} [a typical assumption for Bradley--Terry
  strengths] and say 20 independent complete rank observations,
\cite{hankin2017_rmd} shows that maximum likelihood strengths
identify the correct ranking with a probability of essentially zero.
The situation is worse in cases such as formula 1 motor racing
considered below, which requires us to consider ${\mathcal O}(100)$
competitors.

\cite{hankin2024_hyper3} considers Plackett-Luce likelihood functions
for a race between a competitor of Bradley--Terry strength $a$ and
$n=2$ or $n=3$ cloned competitors, each of strength $b=1-a$.  Here we
consider the generalized case with arbitrary $n$.  Under these
circumstances, an observation is indexed by~$r$, the number of
strength $b$ clones finishing ahead of~$a$, and because we have $n+1$
competitors %%%%%%%%%%%%%%%%%%%%%% $0\leqslant r\leqslant n$.  The
initial field strength is $a+nb=n+(n-1)a$.  The Plackett--Luce
likelihood function~\citep{luce1959,plackett1975} for this situation
would be

@@

\begin{equation}\label{asymptotic}
\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n}
\mathcal{O}\left(n^{-2}\right)
\end{equation}

@@

Previous related work presented the case $n=2$ [in current notation].
Considering the case $n=8$ gives us $8+1=9$ distinct likelihood
functions, Figure~\ref{ninelikes}.  We see an unanticipated asymmetry
between $r$ and $n-r$.  For $r=4$ one might expect that
$\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
performance was median.  However, $\hat{a}\simeq 0.5886$.

@@

One long-standing contention in the athletic world is that of lane
bias~\citep{munro2022}.  Running in the inner lane is frequently held
to be a disadvantage; mechanisms include biomechanical issues arising
from centrifugal acceleration, to competitors' inability to see their
rivals.  Here attention is restricted to heats, for which lane placing
is random.  I consider every heat in Men's 100m Olympic running from
2004--2020 (a total of 41 heats, each of 8 or 9 competitors); only
competitors' ranks are used.  Figure~\ref{plotolymp} shows a
log-likelihood curve for the Bradley--Terry strength of the focal
competitor, here the runner in the innermost lane.  We see that the
evaluate of $0.474$ is very close to the neutral value of $0.5$, for
which the support is only $0.16$, far short of Edwards's two units of
support criterion for rejection~\citep{edwards1972}.  This dataset
contains no evidence to support lane bias.

@@

One might wonder whether the student was better at pure or applied
mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as
the diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.43$, at
${p}_\mathrm{pure}=0.567={p}_\mathrm{applied}=0.713$.  Edwards's two
units of support criterion~\cite{edwards1972} is met, and we are
justified in concluding that the student is indeed more competent at
applied than pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.028$.

>>>>

Ranked data arises whenever multiple entities are sorted by some
criterion; the preferred interpretation is a race in which the
sufficient statistic is the order of competitors crossing the
finishing line.  The Plackett--Luce likelihood
function~\citep{luce1959,plackett1975} is a generalization of the
Bradley--Terry model~\citep{bradley1952} in which non-negative
strengths are assigned to each competitor; without loss of generality,
if the finishing order is $1,2,\ldots,n$ and competitor $i$ has
strength $p_i\geqslant 0$, then the Plackett--Luce likelihood function
will be proportional to

@@

\noindent where the strengths are normalised so $\sum p_i=1$.
However, estimation of the strengths is somewhat difficult if $n$ is
even moderately large: even with as few as 20 competitors following
Zipf's law~\citep{zipf1949} [a typical assumption for Bradley--Terry
  strengths] and say 20 independent complete rank observations,
\cite{hankin2017_rmd} shows that maximum likelihood strengths
identify the correct ranking with a probability of essentially zero.
The situation is worse in cases such as formula 1 motor racing
considered below, which requires us to consider ${\mathcal O}(100)$
competitors.

Consider a Plackett-Luce likelihood functions for a race between a
competitor of Bradley--Terry strength $a$ and $n\geqslant 1$ cloned
competitors, each of strength $b=1-a$.  Under these circumstances, an
observation is indexed by~$r$, the number of strength $b$ clones
finishing ahead of~$a$, and because we have $n+1$
competitors %%%%%%%%%%%%%%%%%%%%%% $0\leqslant r\leqslant n$.  The
initial field strength is $a+nb=n-(n-1)a$.  The Plackett--Luce
likelihood function~\citep{luce1959,plackett1975} for this situation
would be

@@

\begin{equation}\label{asymptotic}
\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n}
+\mathcal{O}\left(n^{-2}\right)
\end{equation}

@@

Previous related work \citep{hankin2024_hyper3} presented the case
$n=2$ [in current notation].  Considering the case $n=8$ gives us
$8+1=9$ distinct likelihood functions, Figure~\ref{ninelikes}.  We see
an unanticipated asymmetry between $r$ and $n-r$.  For $r=4$ one might
expect that $\hat{a}=\frac{1}{2}$ on the grounds that the focal
competitor's performance was median.  However, $\hat{a}\simeq 0.5886$.

@@

One long-standing contention in the athletic world is that of lane
bias~\citep{munro2022}.  Running in the inner lane is frequently held
to be a disadvantage; mechanisms include biomechanical issues arising
from centrifugal acceleration, to competitors' inability to see their
rivals.  Here attention is restricted to heats, for which lane placing
is random.  I consider every heat in Men's 100m Olympic running from
2004--2020 (a total of 41 heats, each of 8 or 9 competitors); only
competitors' ranks are used.  Figure~\ref{plotolymp} shows a
log-likelihood curve for the Bradley--Terry strength of the focal
competitor, here the runner in the innermost lane.  We see that the
evaluate of $0.474$ is very close to the neutral value of $0.5$, for
which the support is only $0.16$, far short of the standard two units
of support criterion for rejection~\citep{edwards1972}.  This dataset
contains no evidence to support lane bias.

@@

One might wonder whether the student was better at pure or applied
mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as
the diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.43$, at
${p}_\mathrm{pure}=0.567={p}_\mathrm{applied}=0.713$.  The two units
of support criterion~\citep{edwards1972} is met, and we are justified
in concluding that the student is indeed more competent at applied
than pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.028$.
